An Empirical Analysis of C Preprocessor Use

### 2000-01-15 macros/papers/emp-use-2.tex
More changes. tbl-subset-categories.tex is automatically generated, but I edited it by hand.

\item Operating on post-processed code, the most common strategy, is
  simple to implement, but the tool's input differs from what the
  programmer sees.  Even when line number mappings are maintained,
  other information is lost in the mapping back to the original source
  code.  Source-level debuggers have no symbolic names or types for
  constants and functions introduced via {\tt \#define}, nor can tools
  trace or set breakpoints in function macros, as they can for
  ordinary functions (even those that have been
  inlined~\cite{Zellweger83:TR}).  An example of a tool working on the
  post-processed code is the use of type inferencing to produce C++
  function templates from C; however, the input ``has been
  preprocessed so that all include files are incorporated and all
  macros expanded''~\cite[p.~145]{Siff-fse96}.  Such preprocessing may
  limit the readability and reusability of the resulting C++
  templates.  Another example is that call graph extractors generally
  work in terms of the post-processed code, even when a human is the
  intended consumer of the call graph~\cite{Murphy-icse18}.

@@

The results reported in this paper omit macros defined only in external
libraries (such as the \file{/usr/include/} hierarchy), even when used in
the package source code; we also omit all macro uses in libraries.  This
prevents library header files and uses of macros defined in them from
swamping the characteristics of the
package code, which is our focus in this study.  The programmer generally
has no control over libraries and their header files, and may not even know
whether a library symbol is defined as a macro.

@@

\item[dangling semicolon]\label{item:swallow-semicolon}
        The macro body takes arguments and expands into a statement or
        multiple statements (39\% of statement macros contain the error).
        Thus, its invocations look like function calls, but it should not
        be used like a function call, as in
\begin{alltt}
    #define ABORT() kill(getpid(),SIGABRT);
    \ldots
    if (*p == 0)
      ABORT();
    else \ldots
\end{alltt}
        because {\tt ABORT();} expands to two statements (the second a null
        statement), which is non-syntactic\,---\,disobeys the language
        grammar\,---\,between the {\tt if} condition and
        {\tt else}.

@@

@@

  %%Numbers: read them off tbl-subset-categories.tex
  %% 2.0%: read off caption of fig/cat-def-frequency.eps
  \captionsmall{Categorization of definitions for each macro name with more
    than one definition.
    For instance, for 869 macro names (28\% of multiply-defined macro
    names), all definitions fall into the expression category, and for 183
    macro names (6.0\% of multiply-defined macro names), all definitions
    are either expressions or constants.
    Rows less than 0.2\%, representing 5 or fewer macro names, are omitted.
    The arrows indicate macros for which the
    names are categorized as ``other'' (4.4\% of all multiply-defined
    macro names; overall, 2.0\% of all macro names are so categorized).
    This chart shows which different macro definition categories tend
    to occur together and assists in understanding the reasons for macro names
    whose multiple definitions cause them to be categorized as ``other''.}

@@

\begin{figure}
% This works, but the figures are upside-down.
% \centerline{\epsfig{file=fig/exp-dep-bymacro.eps,angle=90,height=3.75in}}
% \centerline{\epsfig{file=fig/incl-dep-bymacro.eps,angle=90,height=3.75in}}
% Cannot use ``height'' when rotating by -90 or +270; I do not know why.
% \centerline{\epsfig{file=fig/exp-dep-bymacro.eps,angle=270,width=.5\linewidth}}
% \bigskip
% \centerline{\epsfig{file=fig/incl-dep-bymacro.eps,angle=270,width=.5\linewidth}}
% When Mike exports Excel charts, angle is 270.  When Greg exports, angle is 0.
\centerline{\epsfig{file=fig/exp-dep-bymacro.eps,angle=270,width=.48\linewidth}%
\ %
\epsfig{file=fig/incl-dep-bymacro.eps,angle=270,width=.48\linewidth}}
%%Numbers: read off of Excel exp-dep-bymacro
\captionsmall{Dependences by macro name for \numdependmacronames\ macro
  names in \numdependpackages\ packages.  Each bar represents all macros
  that control at least as many as the labeled percent of the lines in its
  package (but fewer than the next bar).  For instance, the 0.17 bar in the
  expansion dependence chart indicates that 257 macros each control between
  0.17\% and 0.26\% of the entire package that contains that macro.  The
  maximum falls in the penultimate bucket (i.e., the rightmost bucket is
  the first empty one).  The $\varepsilon$ bar represents a small non-zero
  fraction (specifically, $10^{-7}$), so that macros not controlling any lines
  are not conflated with macros controlling very few lines.  A log scale is
  used for the x axis.}

@@

The GNU coding standards~\cite{Stallman97} mention Cpp only to recommend
upper case for macros and use of macros to provide meanings for standard
symbols on platforms that don't support ANSI~C\@.
The Ellemtel coding standards~\cite{ellemtel92} recommend that inline
definitions appear in separate files of their own (not in \file{.h} header
files) that never include other files, that inline functions and {\tt
const} or {\tt enum} constants be used in place of {\tt \#define}, and that
preprocessor macro names be prefixed to avoid name clashes.

@@

Check~\cite{SpulerS92} is a C macro checker that detects some instances of multiple
statements, swallowed {\tt \#else}, side effects, and precedence errors
using largely lexical checks.  Precedence error checks are
performed on macro uses rather than reporting error-prone definitions.  The
authors do not report on the effectiveness of the tool in practice or
justify their tradeoffs between techniques that perform parsing and those
that do not.

@@

Since many of the most problematic uses of Cpp provide portability across
different language dialects or different operating environments,
standardization can obviate many such uses.  Canonicalizing library
function names and calling conventions makes conditional compilation less
necessary (37\% of conditionals test operating system variances) and
incidentally makes all programs more portable, even those that have not
gone to special effort to achieve portability.  This proposal moves the
responsibility for portability (really, conformance to a specification)
from the application program into the library or operating system.

>>>>

\item Operating on post-processed code, the most common strategy, is
  simple to implement, but the tool's input differs from what the
  programmer sees.  Even when line number mappings are maintained,
  other information is lost in the mapping back to the original source
  code.  Source-level debuggers have no symbolic names or types for
  constants and functions introduced via {\tt \#define}, nor can tools
  trace or set breakpoints in function macros, as they can for
  ordinary functions (even those that have been
  inlined~\cite{Zellweger83:TR}).  An example of a tool working on the
  post-processed code is the use of type inferencing to produce C++
  function templates from C; however, the input ``has been
  preprocessed so that all include files are incorporated and all
  macros expanded''~\cite[p.~145]{Siff-fse96}.  Such preprocessing may
  limit the readability of the resulting C++ templates by converting
  terse, high-level, well-named constructs into verbose, low-level code.
  Preprocessing may also limit reusability:  the macro-expanded code will
  perform incorrectly when compiled on a system with different settings
  for the macros.  Another example is that call graph extractors generally
  work in terms of the post-processed code, even when a human is the
  intended consumer of the call graph~\cite{Murphy-icse18}.

@@

The results reported in this paper omit macros defined only in external
libraries (such as the \file{/usr/include/} hierarchy), even when used in
the package source code; we also omit all macro uses in libraries.  
There are many such macros and uses; omitting them
prevents library header files and uses of macros defined in them from
swamping the characteristics of the
package code, which is our focus in this study.  The programmer generally
has no control over libraries and their header files, and may not even know
whether a library symbol is defined as a macro.

@@

\item[dangling semicolon]\label{item:swallow-semicolon}
        The macro body takes arguments and expands into a statement or
        multiple statements (39\% of statement macros contain the error),
        for instance by ending with a semicolon or being enclosed in
        \verb|{|~\ldots~\verb|}|.
        Thus, its invocations look like function calls, but it should not
        be used like a function call, as in
\begin{alltt}
    #define ABORT() kill(getpid(),SIGABRT);
    \ldots
    if (*p == 0)
      ABORT();
    else \ldots
\end{alltt}
        because {\tt ABORT();} expands to two statements (the second a null
        statement), which is non-syntactic\,---\,disobeys the language
        grammar\,---\,between the {\tt if} condition and
        {\tt else}.

@@

        An alternative solution would convert macros whose bodies are
        statements into semicolonless statements (wrapped in {\tt do \verb|{|
        {\rm \ldots}\verb|; }| while (0)}, as noted above).  Invocations of
        such macros look more like function calls are are less error-prone.
        This solution requires notifying users of the change in the macro's
        interface and changing all existing macro uses.

@@

  %%Numbers: read them off tbl-subset-categories.tex
  %% 2.0%: read off caption of fig/cat-def-frequency.eps
  \captionsmall{Categorization of definitions for each macro name with more
    than one definition.
    For instance, for 869 macro names (28\% of multiply-defined macro
    names), all definitions fall into the expression category, and for 183
    macro names (6.0\% of multiply-defined macro names), all definitions
    are either expressions or constants.
    Rows less than 0.2\%, representing 5 or fewer macro names, are omitted.
    The rightmost column indicates the categorization of the macro name; 
    4.4\% of all multiply-defined macro names, and 2.0\% macro names
    overall, are are categorized as ``other''.
    This chart shows which different macro definition categories tend
    to occur together and assists in understanding the reasons for macro names
    whose multiple definitions cause them to be categorized as ``other''.}

@@

\begin{figure}
% This works, but the figures are upside-down.
% \centerline{\epsfig{file=fig/exp-dep-bymacro.eps,angle=90,height=3.75in}}
% \centerline{\epsfig{file=fig/incl-dep-bymacro.eps,angle=90,height=3.75in}}
% Cannot use ``height'' when rotating by -90 or +270; I do not know why.
% \centerline{\epsfig{file=fig/exp-dep-bymacro.eps,angle=270,width=.5\linewidth}}
% \bigskip
% \centerline{\epsfig{file=fig/incl-dep-bymacro.eps,angle=270,width=.5\linewidth}}
% When Mike exports Excel charts, angle is 270.  When Greg exports, angle is 0.
\centerline{\epsfig{file=fig/exp-dep-bymacro.eps,angle=270,width=.48\linewidth}%
\ %
\epsfig{file=fig/incl-dep-bymacro.eps,angle=270,width=.48\linewidth}}
%%Numbers: read off of Excel exp-dep-bymacro
\captionsmall{Dependences by macro name for \numdependmacronames\ macro
  names in \numdependpackages\ packages.  The first bar represents all
  macros that are completely unused in a package.
  Each other bar represents all macros
  that control a specified percentage range of the lines in a
  package.  For instance, the 0.17--0.26 bar in the
  expansion dependence chart indicates that 257 macros each control between
  0.17\% and 0.26\% of the entire package that contains that macro.  The
  maximum falls in the penultimate bucket (i.e., the rightmost bucket is
  the first empty one).  The second bar of each graph represents all macros
  controlling at least one line ($\varepsilon$ is a very small but nonzero
  value), but no more than .01% of the entire package.  A log scale is used
  for the x axis.}

@@

The GNU coding standards~\cite{Stallman97} mention Cpp only to recommend
upper case for macros and use of macros to provide meanings for standard
symbols on platforms that don't support ANSI~C\@.
In any event, most GNU programs do not appear to have been written with
an eye to careful adherence to the standards.
The Ellemtel coding standards~\cite{ellemtel92} recommend that inline
definitions appear in separate files of their own (not in \file{.h} header
files) that never include other files, that inline functions and {\tt
const} or {\tt enum} constants be used in place of {\tt \#define}, and that
preprocessor macro names be prefixed to avoid name clashes.

@@

Check~\cite{SpulerS92} is a C macro checker that detects some instances of multiple
statements, swallowed {\tt \#else}, side effects, and precedence errors
using largely lexical checks.  Precedence error checks are
performed on macro uses rather than reporting error-prone definitions.  The
authors do not report on the effectiveness of the tool in practice, nor do they
justify their tradeoffs between techniques that perform parsing and those
that do not, although such tradeoffs are a major theme of their paper.

@@

Since many of the most problematic uses of Cpp provide portability across
different language dialects or different operating environments,
standardization can obviate many such uses, either in legacy code or, more
easily, in new code.  Canonicalizing library
function names and calling conventions makes conditional compilation less
necessary (37\% of conditionals test operating system variances) and
incidentally makes all programs more portable, even those that have not
gone to special effort to achieve portability.  This proposal moves the
responsibility for portability (really, conformance to a specification)
from the application program into the library or operating system.
